An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-04-18 cal_demo/ms.tex
Addressing some of the easy/straightforward referee comments to cal_demo paper

Direct imaging correlators are a new variety of radio correlator which aim to alleviate both the 
computational strain of forming $\Nant^2$ correlations and the high data throughput associated 
with short timescale science. This is done by performing a spatial fast Fourier transform (FFT) 
to image the antenna voltages, then squaring and averaging in time. This process scales as 
$\mathcal{O}(\Ng \log_2 \Ng)$, where \Ng~is the number of grid points in the FFT \citep{mor11,
 teg09, teg10}. For certain classes of telescopes, significantly those envisioned for next 
 generation cosmology experiments, this scaling is a large improvement over the $\Nant^2$ 
 scaling of traditional methods. Furthermore, because images are generated online, the native 
 output bandwidth will be lowered (assuming $\Ng < \Nant^2$), and has the potential to be 
 lowered even further with online transient processing.

@@

The next step is to grid the antenna signals. MOFF uses the antenna patterns as the gridding 
kernel according to the software holography/A-transpose technique \citep{mor09,bha08}. This 
achieves both an optimal map \citep{teg97b} and places the data on a regular grid. Unlike 
traditional correlators which grid visibilities, the MOFF grids the electric fields directly. This 
operation must be performed for every time step, before any averaging of the data.

After gridding, the imaging portion of the MOFF performs a 2D spatial FFT to form 
instantaneous electric field images of the sky. The pipeline up to this point can be summarized 
through the following equation.
\begin{equation}
\Ethetaest (\s_i,t) = \underbrace{\frac{1}{\Nant} \sum_i e^{2\pi i \ri \cdot \s_i}}_{\mathrm{2D\;FFT}} 
\underbrace{\sum_a \beamr_a(\ri - \ra)}_{\mathrm{Gridding}}
\underbrace{ \frac{g^T_a}{g'_a} \Er{a}^T(t)}_{\mathrm{Calibration}}
\end{equation}
The sum over antennas uses the voltage pattern to grid the calibrated electric fields onto 
regular gridpoint, $i$. The sum over $i$ denotes the 2D FFT to sky coordinates, resulting in an 
estimate for the instantaneous electric field image. We can simplify this expression by 
exchanging the sums to transform the beam term into sky coordinates.
\begin{align}\label{eq:epix}
\Ethetaest(\s_i,t) & = \frac{1}{\Nant} \sum_a \frac{g^T_a}{g'_a}\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra} \nonumber \\
  & \qquad\times \sum_i \beamr_b(\ri-\ra)e^{2\pi i \s_i \cdot (\ri-\ra)} \nonumber\\
& = \frac{1}{\Nant} \sum_a \frac{ g^T_a}{g'_a}\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra}\beamtheta_a(\s_i)
\end{align}
In this form we see that the effect of gridding with the antenna voltage pattern is to attenuate 
the image by a factor of the antenna sky response, $\beamtheta_a(\s_i)$.

@@

The estimated gain amplitudes for the simulations are shown in figure~\ref{fig:sim_amp}. The 
quantity plotted is the magnitude of the estimated gains over the true gains, $\left|g^{(n)}_a/
g^T_a\right|$, which places all antennas on the same scale. We can see the amplitudes 
converge toward their true values around the same time as the phases (iteration $\sim$15). At 
the beginning of calibration we can see the value of the damping factor. At $n=0$, a couple of 
gains are shown to have abnormally high amplitude estimates, notably one about 3.3 times its 
true value (red line). These unbalanced high estimates caused the entire set of gains to be 
under estimated at $n=1$, even with a damping factor of 0.35. By $n=5$ the unbalanced 
amplitudes have been damped out and the calibration continues. Without the damping factor, 
the oscillation seen in the first couple iterations would have been significantly larger and taken 
much longer to fade out.

@@

Finally, we found that seven antennas\footnote{LWA antenna IDs 48, 85, 124, 148, 203, 217, 
and 244 were flagged.} produced unstable gain solutions, and in fact corrupted the entire array. 
We therefore flagged these antennas in our analysis, resulting in a total of 248 antennas to 
calibrate. In each calibration loop, we form $C_{i,\hat{\boldsymbol{s}}_0}, i=1,2,\ldots N_a$ and 
update our gain estimates over 10 timestamps (51.2 ms). We iterate the loop 30 times for a 
total of 1.536 seconds. The results of this calibration experiment are shown in figures~
\ref{fig:data_phase}~--~\ref{fig:data_images}.

@@

The results of our simulations are shown in figure~\ref{fig:errors}. The gain errors from our 
EPICal method is shown with solid lines, while the dashed lines represent the visibility-based 
calibration. The thick lines represent the results of using a full sky model, and the thin lines are 
from the single point source model.  In the case of the full sky model, we see the errors trend 
downward with longer integration times, as expected. The EPICal errors are slightly higher than 
the visibility-based errors, on average about 23\% difference. This is not surprising as EPICal 
only uses the information in a single pixel, while the visibilities use all information from the full 
sky model. However, we see that with this perfect model the errors do behave like noise, and 
the same level of noise can be achieved with a longer integration time (or larger damping 
factor).

>>>>

Direct imaging correlators are a new variety of radio correlator which aim to alleviate both the 
computational strain of forming $\Nant^2$ correlations and the high data throughput associated 
with short timescale science. This is done by performing a spatial fast Fourier transform (FFT) 
to image the antenna voltages, then squaring and averaging in time. This process scales as 
$\mathcal{O}(\Ng \log_2 \Ng)$, where \Ng~is the number of grid points in the FFT \citep{mor11,
 teg09, teg10}. For certain classes of telescopes, significantly those envisioned for next 
 generation cosmology experiments, this scaling is a large improvement over the $\Nant^2$ 
 scaling of traditional methods. Furthermore, because images are generated online, the native 
 output bandwidth will be lowered in cases where $\Ng < \Nant^2$, and has the potential to be 
 lowered even further with online transient processing.

@@

The next step is to grid the antenna signals. MOFF uses the antenna patterns as the gridding 
kernel according to the software holography/A-transpose technique \citep{mor09,bha08}. This 
achieves an optimal map \citep{teg97b} and places the data on a regular grid. Unlike 
traditional correlators which grid visibilities, the MOFF grids the electric fields directly. This 
operation must be performed for every time step, before any averaging of the data.

After gridding, the imaging portion of the MOFF performs a 2D spatial FFT to form 
instantaneous electric field images of the sky. The pipeline up to this point can be summarized 
through the following equation.
\begin{equation}
\Ethetaest (\s_i,t) = \underbrace{\frac{1}{\Nant} \sum_i e^{2\pi i \ri \cdot \s_i}}_{\mathrm{2D\;FFT}} 
\underbrace{\sum_a \beamr_a(\ri - \ra)}_{\mathrm{Gridding}}
\underbrace{ \frac{g^T_a}{g'_a} \Er{a}^T(t)}_{\mathrm{Calibration}}
\end{equation}
The sum over antennas uses the voltage pattern to grid the calibrated electric fields onto 
regular gridpoint, $i$. The sum over $i$ denotes the 2D FFT to sky coordinates, resulting in an 
estimate for the instantaneous electric field image. We can simplify this expression by 
exchanging the sums to transform the beam term into sky coordinates.
\begin{align}\label{eq:epix}
\Ethetaest(\s_i,t) & = \frac{1}{\Nant} \sum_a \frac{g^T_a}{g'_a}\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra} \nonumber \\
  & \qquad\times \sum_i \beamr_a(\ri-\ra)e^{2\pi i \s_i \cdot (\ri-\ra)} \nonumber\\
& = \frac{1}{\Nant} \sum_a \frac{ g^T_a}{g'_a}\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra}\beamtheta_a(\s_i)
\end{align}
In this form we see that the effect of gridding with the antenna voltage pattern is to attenuate 
the image by a factor of the antenna sky response, $\beamtheta_a(\s_i)$.

@@

The estimated gain amplitudes for the simulations are shown in figure~\ref{fig:sim_amp}. The 
quantity plotted is the magnitude of the estimated gains over the true gains, $\left|g^{(n)}_a/
g^T_a\right|$, which places all antennas on the same scale. We can see the amplitudes 
converge toward their true values around the same time as the phases (iteration $\sim$15). At 
the beginning of calibration we can see the importance of the damping factor. At $n=0$, a couple of 
gains are shown to have abnormally high amplitude estimates, notably one about 3.3 times its 
true value (red line). These unbalanced high estimates caused the entire set of gains to be 
under estimated at $n=1$, even with a damping factor of 0.35. By $n=5$ the unbalanced 
amplitudes have been damped out and the calibration continues. Without the damping factor, 
the oscillation seen in the first couple iterations would have been significantly larger and taken 
much longer to fade out.

@@

Finally, we found that seven antennas\footnote{LWA antenna IDs 48, 85, 124, 148, 203, 217, 
and 244 were omitted.} produced unstable gain solutions, and in fact corrupted the entire array. 
We therefore omitted these antennas from our analysis, resulting in a total of 248 antennas to 
calibrate. In each calibration loop, we form $C_{i,\hat{\boldsymbol{s}}_0}, i=1,2,\ldots N_a$ and 
update our gain estimates over 10 timestamps (51.2 ms). We iterate the loop 30 times for a 
total of 1.536 seconds. The results of this calibration experiment are shown in figures~
\ref{fig:data_phase}~--~\ref{fig:data_images}.

@@

The results of our simulations are shown in figure~\ref{fig:errors}. The gain errors from our 
EPICal method are shown with solid lines, while the dashed lines represent the visibility-based 
calibration. The thick lines represent the results of using a full sky model, and the thin lines are 
from the single point source model.  In the case of the full sky model, we see the errors trend 
downward with longer integration times, as expected. The EPICal errors are slightly higher than 
the visibility-based errors, on average about 23\% difference. This is not surprising as EPICal 
only uses the information in a single pixel, while the visibilities use all information from the full 
sky model. However, we see that with this perfect model the errors do behave like noise
in the sense that they integrate down with longer integrations with the expected $t^{-1/2}$ 
trend. 
The same level of gain error as visibility based solutions can be achieved with a longer integration time (or larger damping 
factor).
